---
title: "Habitat *c*omplexity *r*educes feed*i*ng s*t*reng*t*h of fr*e*shwater p*r*edators (CRITTER, Code Repository)"
bibliography: references/CRITTERcode.bib
csl: references/methods-in-ecology-and-evolution.csl
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This code is a supplement to the upcoming publication of Aranbarri et al. (unpublished) investigating the effect of habitat complexity on the feeding functional response of two freshwater invertebrate predators. You will find details on the methodology in the cited publication. We will link to it after we make it publicly available.

**Note that this code is still under construction!**

## License

This code is published under a [**GNU General Public License 3**](https://www.gnu.org/licenses/gpl-3.0.html).

## Authors

-   Björn C. Rall ([0000-0002-3191-8389](https://orcid.org/0000-0002-3191-8389))
    -   [bjoern.rall@uni-konstanz.de](mailto:bjoern.rall@uni-konstanz.de)
    -   Aquatic Ecology and Evolution Group, Limnological Institute, University of Konstanz, Mainaustraße 252, 78464 Konstanz/Egg, Germany
-   Mireia Aranbarri ([0009-0001-3506-0914](https://orcid.org/0009-0001-3506-0914))
    -   [mireia.arambarri@ehu.eus](mailto:mireia.arambarri@ehu.eus)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Lorea Flores ([0000-0002-0082-4072](https://orcid.org/0000-0002-0082-4072))
    -   [lflorescompains@gmail.com](mailto:lflorescompains@gmail.com)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Ioar de Guzmán ([0000-0001-8894-8477](https://orcid.org/0000-0001-8894-8477))
    -   [mirenioar.deguzman@ehu.eus](mailto:mirenioar.deguzman@ehu.eus)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Aitor Larrañaga ([0000-0002-0185-9154](https://orcid.org/0000-0002-0185-9154))
    -   [aitor.larranagaa@ehu.eus](mailto:aitor.larranagaa@ehu.eus)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Julia Reiss ([0000-0002-3740-0046](https://orcid.org/0000-0002-3740-0046))
    -   [julia.reiss@brunel.ac.uk](mailto:julia.reiss@brunel.ac.uk)
    -   Centre for Pollution Research and Policy, Brunel University of London, Uxbridge, UB8 3PH, UK

## Related Works

-   [Data on Zenodo](https://doi.org/10.5281/zenodo.14891980) [@FloresEtAl2025ComplexityReducesFeedingData]

-   [Data on GitHub](https://github.com/b-c-r/CRITTERdata)

-   [R-Code on Zenodo](https://doi.org/10.5281/zenodo.14894598) [@RallEtAl2025ComplexityReducesFeedingCode]

-   [R-Code on GitHub](https://github.com/b-c-r/CRITTERdata)

-   [Statistical Report on Zenodo](https://doi.org/10.5281/zenodo.14898819) [@RallEtAl2025ComplexityReducesFeedingStatistics]

-   [Statistical Report on GitHub](https://github.com/b-c-r/CRITTERstatistics)

-   [Scientific Preprint Article](https://doi.org/10.1101/2025.02.22.639633) [@AranbarriEtAl2025ComplexityReducesFeeding]

## Code Description

We predominantly used functions to organize our R code for this project. Each function is saved in a separate \*.R-file and documented approximately in roxygen2 style [@WickhamEtAl2024Roxygen2InLineDocumentation]. The functions depend mostly hierarchically on each other. All functions are saved in the project's sub folders called functions_*.

### Functions to identify the functional response type

The functions are located in `/functions_type_statistics/`.

#### `phen_type_test`
**`phen_type_test`** is a wrapper around the function `frair_test` [@PritchardEtAl2017FrairToolsFunctional; @PritchardEtAl2017FrairPackageFitting] to test for the type of the functional response using the proportion of prey eaten as response to the initial prey density. If the proportion of prey eaten increases at low prey densities and decreases after reaching a maximum, there is evidence for a type III functional response [@Juliano2001NonlinearCurveFitting]. Proportion data is fitted using a binomial GLM, but see @Crawley2012Book, chapter 16 for an introduction to this topic.

Required packages and their dependencies to be installed:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `purrr` [@WickhamHenry2025PurrrFunctionalProgramming]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `frair` [@PritchardEtAl2017FrairToolsFunctional]

Required packages to be attached:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]


#### `gen_fr_compile`
`gen_fr_compile` compiles the ordinary differential equation (ODE) describing the decay of resource items over time during a feeding functional response trial. The functional response model has a free shape parameter [@Real1977KineticsFunctionalResponse; @Real1979EcologicalDeterminantsFunctional], allowing a continuous shift of the shape of a classic type II functional response [@Holling1959CharacteristicsSimpleTypes] to a type III functional response [@Holling1959ComponentsPredationRevealed]. The feeding rate, $F$, depends on the resource density, $N$, and the parameters of the model are the maximum feeding rate, $F_{max}$, the half saturation density, $N_{half}$, and the shape parameter $q$ [@WilliamsMartinez2004StabilizationChaoticNonpermanent; @Vucic-PesticEtAl2010AllometricFunctionalResponse]:

$$
F = \frac{F_{max}  N^{(1+q)}}{N_{half}^{(1+q)} + N^{(1+q)}}.
$$

The resulting ODE, describing the decay of the resource over time, $dN/dt$, is:

$$
\frac{dN}{dt} =  \frac{-F_{max}  N^{(1+q)}}{N_{half}^{(1+q)} + N^{(1+q)}}.
$$

A numerical simulation is required as there is no analytical solution for this problem [@RosenbaumRall2018FittingFunctionalResponses]. We use the R package “odin” [@FitzJohn2024OdinODEGeneration] to create a fast simulation model in C. This model can be used to estimate the exact shape of the functional response or test if a functional response is type II or type III [@RosenbaumRall2018FittingFunctionalResponses].

Required packages and their dependencies to be installed:

- `odin` [@FitzJohn2024OdinODEGeneration]

Required packages to be attached:

- None

#### `gen_fr_sim`

**`gen_fr_sim`** simulates time series across different initial prey densities. It returns only the number of initial prey items and the number of prey eaten at the end of the time series, mimicking common functional response laboratory experiments. The underlying functional response model is the generalized functional response model [@Real1977KineticsFunctionalResponse; @Real1979EcologicalDeterminantsFunctional]. Note that not integers, but floating numbers are returned by **`gen_fr_sim`**. **`gen_fr_sim`** depends on **`gen_fr_compile`**. Please find details more above.


Required packages and their dependencies to be installed:

- `odin` [@FitzJohn2024OdinODEGeneration]

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_nll`
**`gen_fr_nll`** calculates the negative log likelihood of the generalized functional response model [@Real1977KineticsFunctionalResponse; @Real1979EcologicalDeterminantsFunctional], but see the description of **`gen_fr_compile`** for further information. We calculated the likelihood assuming a binomial distribution, as every prey item has a chance to be eaten or not to be eaten throughout the experimental trial. For further details on the methodology, please read chapter eight of “Ecological models and data in R” [@Bolker2008EcologicalModelsData]. To restrict the fitting to reasonable values of the shape parameter $q$ [@WilliamsMartinez2004StabilizationChaoticNonpermanent; @Vucic-PesticEtAl2010AllometricFunctionalResponse; @RosenbaumRall2018FittingFunctionalResponses] we applied a quadratic penalty on the negative log-likelihood following:

```{r eval=FALSE, include=TRUE}
if(q < q_low){
  nll <- nll + penalty*(q-q_low)^2
} else{
  if(q >= q_up){
    nll <- nll + penalty*(q-q_up)^2
  } else{
    nll <- nll
  }
}
```

`q_low` is set to “0” by default (a type II functional response), `q_up` is set to “1” by default (a “strict” type III functional response), and `penalty` is set to 1000 by default. Especially `q_low` is important, as negative values may lead to an unsolvable time series for `gen_fr_sim` leading to a crash of the fitting process. Even with this restriction, the simulation may fail for extreme values of $F_{max}$ or $N_{half}$; in this case, the function returns `Inf`. Alternative solutions would be that the function returns `NA` [@Bolker2008EcologicalModelsData]. Moreover, the functions require the model parameters $F_{max}$ and $N_{half}$ to be on log10-scale, as this transformation (1) accelerates the fitting procedure and (2) prevents biologically irrelevant negative estimations that would crash the fitting algorithm.

Required packages and their dependencies to be installed:

- `odin` [@FitzJohn2024OdinODEGeneration]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_parms_scan`

 **`gen_fr_parms_scan`** creates Latin hypercube samples for the functional response parameters in a reasonable range and calculates the according negative log-likelihood values. It returns the parameter values with the lowest negative log likelihood of these samples. Non-linear maximum likelihood fitting procedures require starting parameters, generally based on an educated guess [@Bolker2008EcologicalModelsData]. Moreover, these fits may end up in local best fits, and users should re-fit the data using different starting parameters [@Bolker2008EcologicalModelsData]. To overcome manually eyeballing as well as re-shuffling the starting parameters, @JagerAshauer2018ModellingSurvivalChemical suggested creating samples in a reasonable parameter range using and choosing the starting parameters (from the lowest nll value) from these samples. To reduce the number of required samples by keeping the variance of parameter values as wide as possible, it is recommended to use Latin hypercube sampling. `gen_fr_parms_scan` requires the lhs package [@Carnell2024LhsLatinHypercubea]. See also the description of `gen_fr_nll` for further information.

Required packages and their dependencies to be installed:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]
- `odin` [@FitzJohn2024OdinODEGeneration]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_fit`

**`gen_fr_fit`** automatically fits the generalized functional response model [@Real1977KineticsFunctionalResponse; @RosenbaumRall2018FittingFunctionalResponses] to data. In the simplest case, you only need to provide the number of resources eaten (count data) and the initial resource density (count data): the function does the rest, including initial parameter value guessing. See the parameters section and the code example for more options. If your experiment ran a day, but you want to have the maximum feeding rate on an hourly basis, you can enter t_end = 24. See also the description of `gen_fr_nll` and `gen_fr_parms_scan` for further information.

Required packages and their dependencies to be installed:

- `bbmle` [@BolkerEtAl2023BbmleToolsGeneral]
- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]
- `odin` [@FitzJohn2024OdinODEGeneration]

Required packages to be attached:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_fit_all`

**`gen_fr_fit_all`** fits the generalized functional response model [@Real1977KineticsFunctionalResponse; @RosenbaumRall2018FittingFunctionalResponses] by running `gen_fr_fit` for all treatments in parallel.

Required packages and their dependencies to be installed:

- `bbmle` [@BolkerEtAl2023BbmleToolsGeneral]
- `doParallel` [@MicrosoftWeston2022DoParallelForeachParallel]
- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]
- `odin` [@FitzJohn2024OdinODEGeneration]
- `purrr` [@WickhamHenry2025PurrrFunctionalProgramming]

Required packages to be attached:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

### Habitat Complexity Statistics Functions

#### `rrpe_sim`

***Description***:

`rrpe_sim` simulates the type II functional response [@Holling1959CharacteristicsSimpleTypes; @Real1977KineticsFunctionalResponse]. The feeding rate, $F$, is determined by the model parameters handling time, $T_{h}$, and attack rate, $a$ [@Holling1959CharacteristicsSimpleTypes] or maximum feeding rate, $F_{max}$, and half saturation density, $N_{half}$ [@Real1977KineticsFunctionalResponse]:

$$
F = \frac{a N}{1 + a T_h N} = \frac{F_{max} N}{N_{half} + N},
$$

where $N$ is the current resource density. These equations, however, require a constant resource density, which is not given in most functional response experiments. To take the temporal decline of the resource into account, we apply Rogers' Random Equation [@Royama1971ComparativeStudyModels; @Rogers1972RandomSearchInsect]:

$$
N_{eaten} = N_{initial} (1 - e^{a (N_{eaten}T_{h} - P  T_{end})}) = N_{initial} (1 - e^{\frac{N_{eaten} - F_{max} P  T_{end}}{N_{half}}}),
$$

where N_eaten are the eaten resource items, N_initial is the initial resource density, P is the predator density, and T_end is the experimental duration (time). These equations contain N_eaten on both sides and are only solvable iteratively [@Juliano2001NonlinearCurveFitting; @VoneshBolker2005CompensatoryLarvalResponses]. Alternatively, @Bolker2008EcologicalModelsData solved this issue by using the Lambert $W$
function [@CorlessEtAl1996LambertWFunction]:

$$
N_{eaten} = N_{initial} - \frac{W(a T_h N_{initial} e^{a(T_h N_{initial} - P T_{end})})}{a T_h}  = N_{initial} - N_{half} W(\frac{N_{initial}}{N_{half}}  e^{-\frac{F_{max} P T_{end}- N_{initial}}{N_{half}}})
$$

We apply these functions to compute the type II functional response.

***Parameters***:

-   `fr_style` [string]: either "Holling" or "Real".
-   `n_initial` [numeric]: a vector of initial prey densities (can also be a single value).
-   `p` [numeric]: a single value of a fixed predator density. The default value is 1.
-   `a` [numeric]: the attack rate, a single value. Only for Holling-style.
-   `t_h` [numeric]: the handling time, a single value. Only for Holling-style.
-   `f_max` [numeric]: the maximum feeding rate, a single value. Only for Real-style.
-   `n_half` [numeric]: the half saturation density, a single value. Only for Real-style.
-   `t_end` [numeric]: the time were the feeding ends. A single value; default = 1 (e.g. 1 day).


***Required packages***:

- `emdbook` [@Bolker2023EmdbookSupportFunctions]


#### `rrpe_nll_mod01r` to `rrpe_nll_mod16r` and `rrpe_nll_mod01h` to `rrpe_nll_mod16h`

***Description***:

All functions from `rrpe_nll_mod01r` to `rrpe_nll_mod16r` and `rrpe_nll_mod01h` to `rrpe_nll_mod16h` calculate the negative log likelihood of using experimental functional response data (eaten resource items as a function of resource density) using the Michaelis-Menten Type II functional response [@Real1977KineticsFunctionalResponse] from the `rrpe_sim` function. Functions having `r` in their name return the parameters for the Real-style functional response [@Real1977KineticsFunctionalResponse]: $F_{max}$ and $N_{half}$. Functions having `h` in their name return the parameters for the Holling-style functional response [@Holling1959CharacteristicsSimpleTypes]: $T_{handling}$ and $a$.
We calculated the likelihood by assuming a binomial distribution of the depended data, i.e., whether a resource item can be eaten or not at the end of the experimental trial. See @Bolker2008EcologicalModelsData, chapter 8 for details. Moreover, the functions require the model parameters on log-scale, as this transformation (1) accelerates the fitting procedure and (2) prevents biologically irrelevant negative estimations that would crash the fitting algorithm.

***Parameters***:

-   `n_eaten` [numeric]: a data vector of initial prey densities. **MUST BE INTEGERS (i.e. 0, 1, 2, ...)!**
-   `n_initial` [numeric]: a data vector initial prey densities. **SHOULD BE INTEGERS (i.e. 0, 1, 2, ...)! MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `n_rings` [numeric]: a data vector of number of rings provided as habitat structure. **MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `complexity` [numeric]: a data vector of level of complexity. **MUST BE ON OF THE FOLLOWING: 0, 1, 2, 3, 4! MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `p` [numeric]: a single value of a fixed predator density. The default value is 1. **SHOULD BE INTEGERS (i.e. 0, 1, 2, ...)!**
-   `a_log10` [numeric]: the $log_{10}$ of the attack rate across treatments, a single value. Only for Holling-style functions and if habitat has no effect on attack rate.
-   `a_hab0_log10` [numeric]: the $log_{10}$ of the attack rate for habitat being absent, a single value. Only for Holling-style functions and if habitat presence is considered for attack rate.
-   `a_hab1_log10` [numeric]: the $log_{10}$ of the attack rate for habitat being present, a single value. Only for Holling-style functions and if habitat presence is considered for attack rate.
-   `a_0_log10` [numeric]: the $log_{10}$ of the attack rate for complexity level 0, a single value. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_1_log10` [numeric]: the $log_{10}$ of the attack rate for complexity level 1, a single value. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_2_log10` [numeric]: the $log_{10}$ of the attack rate for complexity level 2, a single value. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_3_log10` [numeric]: the $log_{10}$ of the attack rate for complexity level 3, a single value. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_4_log10` [numeric]: the $log_{10}$ of the attack rate for complexity level 4, a single value. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_intercept_log10` [numeric]: the $log_{10}$ of the intercept of the attack rate, a single value. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for attack rate.
-   `a_slope` [numeric]: the slope of the attack rate, a single value. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for attack rate.
-   `t_h_log10` [numeric]: the $log_{10}$ of the handling time across treatments, a single value. Only for Holling-style functions and if habitat has no effect on handling time.
-   `t_h_hab0_log10` [numeric]: the $log_{10}$ of the handling time for habitat being absent, a single value. Only for Holling-style functions and if habitat presence is considered for handling time.
-   `t_h_hab1_log10` [numeric]: the $log_{10}$ of the handling time for habitat being present, a single value. Only for Holling-style functions and if habitat presence is considered for handling time.
-   `t_h_0_log10` [numeric]: the $log_{10}$ of the handling time for complexity level 0, a single value. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_1_log10` [numeric]: the $log_{10}$ of the handling time for complexity level 1, a single value. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_2_log10` [numeric]: the $log_{10}$ of the handling time for complexity level 2, a single value. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_3_log10` [numeric]: the $log_{10}$ of the handling time for complexity level 3, a single value. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_4_log10` [numeric]: the $log_{10}$ of the handling time for complexity level 4, a single value. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_intercept_log10` [numeric]: the $log_{10}$ of the intercept of the handling time, a single value. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for handling time.
-   `t_h_slope` [numeric]: the slope of the handling time, a single value. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for handling time.
-   `n_half_log10` [numeric]: the $log_{10}$ of the half saturation density across treatments, a single value. Only for Real-style functions and if habitat has no effect on half saturation density.
-   `n_half_hab0_log10` [numeric]: the $log_{10}$ of the half saturation density for habitat being absent, a single value. Only for Real-style functions and if habitat presence is considered for half saturation density.
-   `n_half_hab1_log10` [numeric]: the $log_{10}$ of the half saturation density for habitat being present, a single value. Only for Real-style functions and if habitat presence is considered for half saturation density.
-   `n_half_0_log10` [numeric]: the $log_{10}$ of the half saturation density for complexity level 0, a single value. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_1_log10` [numeric]: the $log_{10}$ of the half saturation density for complexity level 1, a single value. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_2_log10` [numeric]: the $log_{10}$ of the half saturation density for complexity level 2, a single value. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_3_log10` [numeric]: the $log_{10}$ of the half saturation density for complexity level 3, a single value. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_4_log10` [numeric]: the $log_{10}$ of the half saturation density for complexity level 4, a single value. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_intercept_log10` [numeric]: the $log_{10}$ of the intercept of the half saturation density, a single value. Only for Real-style functions and if the amount of habitat (number of rings) is considered for half saturation density.
-   `n_half_slope` [numeric]: the slope of the half saturation density, a single value. Only for Real-style functions and if the amount of habitat (number of rings) is considered for half saturation density.
-   `f_max_log10` [numeric]: the $log_{10}$ of the maximum feeding rate across treatments, a single value. Only for Real-style functions and if habitat has no effect on maximum feeding rate.
-   `f_max_hab0_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for habitat being absent, a single value. Only for Real-style functions and if habitat presence is considered for maximum feeding rate.
-   `f_max_hab1_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for habitat being present, a single value. Only for Real-style functions and if habitat presence is considered for maximum feeding rate.
-   `f_max_0_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 0, a single value. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_1_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 1, a single value. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_2_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 2, a single value. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_3_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 3, a single value. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_4_log10` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 4, a single value. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_intercept_log10` [numeric]: the $log_{10}$ of the intercept of the maximum feeding rate, a single value. Only for Real-style functions and if the amount of habitat (number of rings) is considered for maximum feeding rate.
-   `f_max_slope` [numeric]: the slope of the maximum feeding rate, a single value. Only for Real-style functions and if the amount of habitat (number of rings) is considered for maximum feeding rate.
-   `t_end` [numeric]: the time were the feeding ends. A single value; default = 1 (e.g. 1 day).


***Required packages***:

- `emdbook` [@Bolker2023EmdbookSupportFunctions; @Bolker2008EcologicalModelsData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `rrpe_parms_scan_mod01r` to `rrpe_parms_scan_mod16r` and `rrpe_parms_scan_mod01h` to `rrpe_parms_scan_mod16h`

***Description***:

All `rrpe_parms_scan_mod*` functions create Latin hypercube samples for the functional response parameters in a reasonable range and calculate the according negative log likelihood values. They return the parameter values with the lowest negative log likelihood of these samples. Non-linear maximum likelihood fitting procedures require starting parameters, generally based on an educated guess [@Bolker2008EcologicalModelsData]. Moreover, these fits may end up in local best fits, and users should re-fit the data using different starting parameters [@Bolker2008EcologicalModelsData]. To overcome manually eyeballing as well as re-shuffling the starting parameters, @JagerAshauer2018ModellingSurvivalChemical suggested creating samples in a reasonable parameter range using and choosing the starting parameters (from the lowest nll value) from these samples. To reduce the number of required samples by keeping the variance of parameter values as wide as possible, it is recommended to use Latin hypercube sampling. `rrpe_parms_scan_mod*16r*` require the lhs package [@Carnell2024LhsLatinHypercubea]. 

***Parameters***:

-   `n_eaten` [numeric]: a data vector of initial prey densities. **MUST BE INTEGERS (i.e. 0, 1, 2, ...)!**
-   `n_initial` [numeric]: a data vector initial prey densities. **SHOULD BE INTEGERS (i.e. 0, 1, 2, ...)! MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `n_rings` [numeric]: a data vector of number of rings provided as habitat structure. **MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `complexity` [numeric]: a data vector of level of complexity. **MUST BE ON OF THE FOLLOWING: 0, 1, 2, 3, 4! MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `p` [numeric]: a single value of a fixed predator density. The default value is 1. **SHOULD BE INTEGERS (i.e. 0, 1, 2, ...)!**
-   `a_log10_range` [numeric]: the $log_{10}$ of the attack rate across treatments, **two values**. Only for Holling-style functions and if habitat has no effect on attack rate.
-   `a_hab0_log10_range` [numeric]: the $log_{10}$ of the attack rate for habitat being absent, **two values**. Only for Holling-style functions and if habitat presence is considered for attack rate.
-   `a_hab1_log10_range` [numeric]: the $log_{10}$ of the attack rate for habitat being present, **two values**. Only for Holling-style functions and if habitat presence is considered for attack rate.
-   `a_0_log10_range` [numeric]: the $log_{10}$ of the attack rate for complexity level 0, **two values**. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_1_log10_range` [numeric]: the $log_{10}$ of the attack rate for complexity level 1, **two values**. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_2_log10_range` [numeric]: the $log_{10}$ of the attack rate for complexity level 2, **two values**. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_3_log10_range` [numeric]: the $log_{10}$ of the attack rate for complexity level 3, **two values**. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_4_log10_range` [numeric]: the $log_{10}$ of the attack rate for complexity level 4, **two values**. Only for Holling-style functions and if complexity is considered for attack rate.
-   `a_intercept_log10_range` [numeric]: the $log_{10}$ of the intercept of the attack rate, **two values**. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for attack rate.
-   `a_slope_range` [numeric]: the slope of the attack rate, **two values**. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for attack rate.
-   `t_h_log10_range` [numeric]: the $log_{10}$ of the handling time across treatments, **two values**. Only for Holling-style functions and if habitat has no effect on handling time.
-   `t_h_hab0_log10_range` [numeric]: the $log_{10}$ of the handling time for habitat being absent, **two values**. Only for Holling-style functions and if habitat presence is considered for handling time.
-   `t_h_hab1_log10_range` [numeric]: the $log_{10}$ of the handling time for habitat being present, **two values**. Only for Holling-style functions and if habitat presence is considered for handling time.
-   `t_h_0_log10_range` [numeric]: the $log_{10}$ of the handling time for complexity level 0, **two values**. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_1_log10_range` [numeric]: the $log_{10}$ of the handling time for complexity level 1, **two values**. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_2_log10_range` [numeric]: the $log_{10}$ of the handling time for complexity level 2, **two values**. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_3_log10_range` [numeric]: the $log_{10}$ of the handling time for complexity level 3, **two values**. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_4_log10_range` [numeric]: the $log_{10}$ of the handling time for complexity level 4, **two values**. Only for Holling-style functions and if complexity is considered for handling time.
-   `t_h_intercept_log10_range` [numeric]: the $log_{10}$ of the intercept of the handling time, **two values**. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for handling time.
-   `t_h_slope_range` [numeric]: the slope of the handling time, **two values**. Only for Holling-style functions and if the amount of habitat (number of rings) is considered for handling time.
-   `n_half_log10_range` [numeric]: the $log_{10}$ of the half saturation density across treatments, **two values**. Only for Real-style functions and if habitat has no effect on half saturation density.
-   `n_half_hab0_log10_range` [numeric]: the $log_{10}$ of the half saturation density for habitat being absent, **two values**. Only for Real-style functions and if habitat presence is considered for half saturation density.
-   `n_half_hab1_log10_range` [numeric]: the $log_{10}$ of the half saturation density for habitat being present, **two values**. Only for Real-style functions and if habitat presence is considered for half saturation density.
-   `n_half_0_log10_range` [numeric]: the $log_{10}$ of the half saturation density for complexity level 0, **two values**. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_1_log10_range` [numeric]: the $log_{10}$ of the half saturation density for complexity level 1, **two values**. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_2_log10_range` [numeric]: the $log_{10}$ of the half saturation density for complexity level 2, **two values**. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_3_log10_range` [numeric]: the $log_{10}$ of the half saturation density for complexity level 3, **two values**. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_4_log10_range` [numeric]: the $log_{10}$ of the half saturation density for complexity level 4, **two values**. Only for Real-style functions and if complexity is considered for half saturation density.
-   `n_half_intercept_log10_range` [numeric]: the $log_{10}$ of the intercept of the half saturation density, **two values**. Only for Real-style functions and if the amount of habitat (number of rings) is considered for half saturation density.
-   `n_half_slope_range` [numeric]: the slope of the half saturation density, **two values**. Only for Real-style functions and if the amount of habitat (number of rings) is considered for half saturation density.
-   `f_max_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate across treatments, **two values**. Only for Real-style functions and if habitat has no effect on maximum feeding rate.
-   `f_max_hab0_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for habitat being absent, **two values**. Only for Real-style functions and if habitat presence is considered for maximum feeding rate.
-   `f_max_hab1_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for habitat being present, **two values**. Only for Real-style functions and if habitat presence is considered for maximum feeding rate.
-   `f_max_0_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 0, **two values**. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_1_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 1, **two values**. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_2_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 2, **two values**. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_3_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 3, **two values**. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_4_log10_range` [numeric]: the $log_{10}$ of the maximum feeding rate for complexity level 4, **two values**. Only for Real-style functions and if complexity is considered for maximum feeding rate.
-   `f_max_intercept_log10_range` [numeric]: the $log_{10}$ of the intercept of the maximum feeding rate, **two values**. Only for Real-style functions and if the amount of habitat (number of rings) is considered for maximum feeding rate.
-   `f_max_slope_range` [numeric]: the slope of the maximum feeding rate, **two values**. Only for Real-style functions and if the amount of habitat (number of rings) is considered for maximum feeding rate.
-   `t_end` [numeric]: the time were the feeding ends. A single value; default = 1 (e.g. 1 day).
-   `no_lhs_samples` [numeric]: a single integer value; the number of random latin hypercube samplings. **MUST BE AN INTEGER, E.G. 100, 1000, 2000, ...**. Default = 1000.

***Required packages***:

- `emdbook` [@Bolker2023EmdbookSupportFunctions; @Bolker2008EcologicalModelsData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]


#### `rrpe_fit_mod01r` to `rrpe_fit_mod16r` and `rrpe_fit_mod01h` to `rrpe_fit_mod16h`

***Description***:

All functions from `rrpe_fit_mod01r` to `rrpe_fit_mod16r` and `rrpe_fit_mod01h` to `rrpe_fit_mod16h` allow for fitting the Rogers' Random Predator Equation function [@Royama1971ComparativeStudyModels; @Rogers1972RandomSearchInsect] using the Michaelis-Menten version of the Type II functional response [@Real1977KineticsFunctionalResponse]. Functions having `r` in their name return the parameters for the Real-style functional response [@Real1977KineticsFunctionalResponse]: $F_{max}$ and $N_{half}$. Functions having `h` in their name return the parameters for the Holling-style functional response [@Holling1959CharacteristicsSimpleTypes]: $T_{handling}$ and $a$. You only need to provide the number of resources eaten and the initial resource density. The function does the rest. But see the parameters section and the example for more possibilities. E.g., if your trials ran a day, but you want to have the maximum feeding rate on an hourly basis, you can enter t_end = 24.

***Parameters***:

-   `n_eaten` [numeric]: a data vector of initial prey densities. **MUST BE INTEGERS (i.e. 0, 1, 2, ...)!**
-   `n_initial` [numeric]: a data vector initial prey densities. **SHOULD BE INTEGERS (i.e. 0, 1, 2, ...)! MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `n_rings` [numeric]: a data vector of number of rings provided as habitat structure. **MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `complexity` [numeric]: a data vector of level of complexity. **MUST BE ON OF THE FOLLOWING: 0, 1, 2, 3, 4! MUST BE OF SAME LENGTH AS `n_eaten`!**
-   `p` [numeric]: a single value of a fixed predator density. The default value is 1. **SHOULD BE INTEGERS (i.e. 0, 1, 2, ...)!**
-   `t_end` [numeric]: the time were the feeding ends. A single value; default = 1 (e.g. 1 day).
-   `no_lhs_samples` [numeric]: a single integer value; the number of random latin hypercube samplings. **MUST BE AN INTEGER, E.G. 100, 1000, 2000, ...**. Default = 1000.
-   `range_multiplier` [numeric]: the multipliers with which the current best parameters should be multiplied for the validation random latin hypercube sampling.
-   `rel_f_max_range` [numeric]: these two values are multiplied by the largest feeding value of n_eaten to set the initial boundaries to seek for a reasonable starting value of f_max, default = c(0.6, 0.95)
-   `rel_n_half_range` [numeric]: these two values are multiplied by the largest starting density value, n_initial, to set the initial boundaries to seek for a reasonable starting value of n_half, default = c(0.2, 0.8)
-   `slope_range` [numeric]: the range of initial slopes tested for the log-linear relationship with number of rings. It applies for any FR variable if fitted to number of rings. The intercept is calculated using complexity level = 0. Default is c(-0.05, 0.05).
-   `witer_max` [numeric]: how many fits should be performed without convergence?
-   `mle2_tol` [numeric]: the tolerance of a single mle2 fit.
-   `val_tol` [numeric]: the tolerance of the validation.
-   `set_seed` [numeric]: set seed for better reproducibility? default = TRUE.
-   `seed_value` [numeric]: seed value, default = 123.

***Required packages***:

- `bbmle` [@BolkerEtAl2023BbmleToolsGeneral; @Bolker2008EcologicalModelsData]
- `emdbook` [@Bolker2023EmdbookSupportFunctions; @Bolker2008EcologicalModelsData]
- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]

### Functions for creating a pdf report using Rmarkdown

The functions described below are especially programmed for the purpose of our statistical report [@RallEtAl2025ComplexityReducesFeedingStatistics] and our scienitif publication [@AranbarriEtAl2025ComplexityReducesFeeding]. They are likely only usable in this context.

#### `phen_type_table`
**`phen_type_table`** takes a `phen_type_test` output and creates a nice table for the statistical report. Note that the function uses options from `kableEXTRA` [@Zhu2024KableExtraConstructComplex] that will only work for LaTeX/PDF outputs. This function is rather hard-coded and only useful in the CRITTER project.

Required packages and their dependencies to be installed:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `kableExtra` [@Zhu2024KableExtraConstructComplex]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_table`

**`gen_fr_table`** creates a nice-looking table using the output from `gen_fr_fit_all`. Hard-coded and only useful in this project.

Required packages and their dependencies to be installed:

- `bbmle` [@BolkerEtAl2023BbmleToolsGeneral]
- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `kableExtra` [@Zhu2024KableExtraConstructComplex]
- `knitr` [@Xie2024KnitrGeneralpurposePackage]

Required packages to be attached:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `create_h_table`

**`create_h_table`** creates a nice-looking table using the output from all fits using functions `rrpe_fit_mod01r` to `rrpe_fit_mod16r` and `rrpe_fit_mod01h` to `rrpe_fit_mod16h`. Hard-coded and only useful in this project.

***Parameters***:
-   `h_test_results`: a list of mle2 objects created by running all tests from `rrpe_fit_mod01r` to `rrpe_fit_mod16r`.
-   `cut_after`: the number of results that should be displayed.
-   `caption_text`: the tables caption text.

***Required packages***:

- `bbmle` [@BolkerEtAl2023BbmleToolsGeneral; @Bolker2008EcologicalModelsData]
- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `kableExtra` [@Zhu2024KableExtraConstructComplex]
- `knitr` [@Xie2024KnitrGeneralpurposePackage]


#### `create_summary_table`

**`create_summary_table`** creates a nice-looking table using a mle2 object from a fit of one of the models: `rrpe_fit_mod01r` to `rrpe_fit_mod16r` and `rrpe_fit_mod01h` to `rrpe_fit_mod16h`. In addition, it also calculates the confidence intervals using the *population prediction intervals* (see @Bolker2008EcologicalModelsData, chapter 7.5.3) based on the Hard-coded and only useful in this project.

***Parameters***:
-   `h_test_results`: a list of mle2 objects created by running all tests from `rrpe_fit_mod01r` to `rrpe_fit_mod16r`.
-   `ci_reps` [integer]: the number of CI samples that should be taken, default = 10000.
-   `ci_levels` [numeric]: a vector of two numbers, the lower and upper CI levels. Default is `c(0.025, 0.975)`.
-   `dec_places`: the number of decimal places that should be displayed. Default is 3.
-   `par_names`: a vector of parameter names that should be displayed. Default is `c($T_{h}$, $a$)`.
-   `unlog`: the fit was done using parameters on log-scale (not the slope). Should these be displayed on regular scale? Default is `c(TRUE, TRUE)`. **Note, that you must add `FALSE` if the parameter was a slope**.
-   `caption_text`: the tables caption text.


***Required packages***:

- `bbmle` [@BolkerEtAl2023BbmleToolsGeneral; @Bolker2008EcologicalModelsData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `kableExtra` [@Zhu2024KableExtraConstructComplex]
- `knitr` [@Xie2024KnitrGeneralpurposePackage]
- `MASS` [@RipleyEtAl2025MASSSupportFunctions]




## Funding Information

-   Mireia Aranbarri was funded by the **Investigo Programm funded by the NextGenerationEU initiative**.
-   Lorea Flores was funded by a grant by the **Spanish Ministry of Education and Culture**.
-   Ioar de Guzmán was funded by the **Spanish Ministry of Science, Innovation and Universities (TED2021-129966B-C31)**.
-   Julia Reiss was supported by a **Royal Society of London Starting Grant**.
-   Björn C. Rall gratefully acknowledges the funding by the **German Science Foundation (DFG) to the Research Unit DynaSym (FOR 5064)**.

## References
