---
title: "*C*omplexity *r*educes feed*i*ng s*t*reng*t*h of fr*e*shwater p*r*edators (CRITTER, Code Repository)"
bibliography: references/CRITTERcode.bib
csl: references/methods-in-ecology-and-evolution.csl
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This code is a supplement to the upcoming publication of Aranbarri et al. (unpublished) investigating the effect of habitat complexity on the feeding functional response of two freshwater invertebrate predators. You will find details on the methodology in the cited publication. We will link to it after we make it publicly available.

**Note that this code is still under construction!**

## License

This code is published under a [**GNU General Public License 3**](https://www.gnu.org/licenses/gpl-3.0.html).

## Authors

-   Björn C. Rall ([0000-0002-3191-8389](https://orcid.org/0000-0002-3191-8389))
    -   [bjoern.rall@uni-konstanz.de](bjoern.rall@uni-konstanz.de)
    -   Aquatic Ecology and Evolution Group, Limnological Institute, University of Konstanz, Mainaustraße 252, 78464 Konstanz/Egg, Germany
-   Mireia Aranbarri ([0009-0001-3506-0914](https://orcid.org/0009-0001-3506-0914))
    -   [mireia.arambarri@ehu.eus](mireia.arambarri@ehu.eus)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Lorea Flores ([0000-0002-0082-4072](https://orcid.org/0000-0002-0082-4072))
    -   [lflorescompains@gmail.com](lflorescompains@gmail.com)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Ioar de Guzmán ([0000-0001-8894-8477](https://orcid.org/0000-0001-8894-8477))
    -   [mirenioar.deguzman@ehu.eus](mirenioar.deguzman@ehu.eus)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Aitor Larrañaga ([0000-0002-0185-9154](https://orcid.org/0000-0002-0185-9154))
    -   [aitor.larranagaa@ehu.eus](aitor.larranagaa@ehu.eus)
    -   Laboratory of Stream Ecology, Department of Plant Biology and Ecology, Faculty of Science and Technology, University of the Basque Country, UPV/EHU PO Box 644, 48080 Bilbao, Spain
-   Julia Reiss ([0000-0002-3740-0046](https://orcid.org/0000-0002-3740-0046))
    -   [julia.reiss@brunel.ac.uk](julia.reiss@brunel.ac.uk)
    -   Division of Environmental Sciences, College of Health, Medicine and Life Sciences, Brunel University of London, Uxbridge, UB8 3PH, UK

## Related Works

-   [Data on Zenodo](https://doi.org/10.5281/zenodo.14891980) [@FloresEtAl2025ComplexityReducesFeedingData]

-   [Data on GitHub](https://github.com/b-c-r/CRITTERdata)

-   [R-Code on Zenodo](https://doi.org/10.5281/zenodo.14894598) [@RallEtAl2025ComplexityReducesFeedingCode]

-   [R-Code on GitHub](https://github.com/b-c-r/CRITTERdata)

-   [Statistical Report on GitHub](https://github.com/b-c-r/CRITTERstatistics)

-   [Statistical Report on Zenodo](https://doi.org/10.5281/zenodo.14898819) [@RallEtAl2025ComplexityReducesFeedingStatistics]

-   Scientific Preprint Paper (link tba)

## Code Description

We predominantly used functions to organize our R code for this project. Each function is saved in a separate \*.R-file and documented approximately in roxygen2 style [@WickhamEtAl2024Roxygen2InLineDocumentation]. The functions depend mostly hierarchically on each other. All functions are saved in the project's sub folder called functions_*.

### Phenomenological Functional Response Type Test

The functions are located in `/functions_phen_test/`.

#### `phen_type_test`
**`phen_type_test`** is a wrapper around the function `frair_test` [@PritchardEtAl2017FrairToolsFunctional; @PritchardEtAl2017FrairPackageFitting] to test for the type of the functional response using the proportion of prey eaten as response to the initial prey density. If the proportion of prey eaten increases at low prey densities and decreases after reaching a maximum, there is evidence for a type III functional response [@Juliano2001NonlinearCurveFitting]. Proportion data is fitted using a binomial GLM, but see @Crawley2012Book, chapter 16 for an introduction to this topic.

Required packages and their dependencies to be installed:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `purrr` [@WickhamHenry2025PurrrFunctionalProgramming]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `frair` [@PritchardEtAl2017FrairToolsFunctional]

Required packages to be attached:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `phen_type_table`
**`phen_type_table`** takes a `phen_type_test` output and creates a nice table for the statistical report. Note that the function uses options from `kableEXTRA` [@Zhu2024KableExtraConstructComplex] that will only work for LaTeX/PDF outputs. This function is rather hard-coded and only useful in the CRITTER project.

Required packages and their dependencies to be installed:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `kableExtra` [@Zhu2024KableExtraConstructComplex]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

### Generalized Functional Response Model Functions

The functions are located in `/functions_gen_fr/`.

#### `gen_fr_compile`
`gen_fr_compile` compiles the ordinary differential equation (ODE) describing the decay of resource items over time during a feeding functional response trial. The functional response model has a free shape parameter [@Real1977KineticsFunctionalResponse; @Real1979EcologicalDeterminantsFunctional], allowing a continuous shift of the shape of a classic type II functional response [@Holling1959CharacteristicsSimpleTypes] to a type III functional response [@Holling1959ComponentsPredationRevealed]. The feeding rate, $F$, depends on the resource density, $N$, and the parameters of the model are the maximum feeding rate, $F_{max}$, the half saturation density, $N_{half}$, and the shape parameter $q$ [@WilliamsMartinez2004StabilizationChaoticNonpermanent; @Vucic-PesticEtAl2010AllometricFunctionalResponse]:

$$
F = \frac{F_{max}  N^{(1+q)}}{N_{half}^{(1+q)} + N^{(1+q)}}.
$$

The resulting ODE, describing the decay of the resource over time, $dN/dt$, is:

$$
\frac{dN}{dt} =  \frac{-F_{max}  N^{(1+q)}}{N_{half}^{(1+q)} + N^{(1+q)}}.
$$

A numerical simulation is required as there is no analytical solution for this problem [@RosenbaumRall2018FittingFunctionalResponses]. We use the R package “odin” [@FitzJohn2024OdinODEGeneration] to create a fast simulation model in C. This model can be used to estimate the exact shape of the functional response or test if a functional response is type II or type III [@RosenbaumRall2018FittingFunctionalResponses].

Required packages and their dependencies to be installed:

- `odin` [@FitzJohn2024OdinODEGeneration]

Required packages to be attached:

- None

#### `gen_fr_sim`

**`gen_fr_sim`** simulates time series across different initial prey densities. It returns only the number of initial prey items and the number of prey eaten at the end of the time series, mimicking common functional response laboratory experiments. The underlying functional response model is the generalized functional response model [@Real1977KineticsFunctionalResponse; @Real1979EcologicalDeterminantsFunctional]. Note that not integers, but floating numbers are returned by **`gen_fr_sim`**. **`gen_fr_sim`** depends on **`gen_fr_compile`**. Please find details more above.


Required packages and their dependencies to be installed:

- `odin` [@FitzJohn2024OdinODEGeneration]

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_nll`
**`gen_fr_nll`** calculates the negative log likelihood of the generalized functional response model [@Real1977KineticsFunctionalResponse; @Real1979EcologicalDeterminantsFunctional], but see the description of **`gen_fr_compile`** for further information. We calculated the likelihood assuming a binomial distribution, as every prey item has a chance to be eaten or not to be eaten throughout the experimental trial. For further details on the methodology, please read chapter eight of “Ecological models and data in R” [@Bolker2008EcologicalModelsData]. To restrict the fitting to reasonable values of the shape parameter $q$ [@WilliamsMartinez2004StabilizationChaoticNonpermanent; @Vucic-PesticEtAl2010AllometricFunctionalResponse; @RosenbaumRall2018FittingFunctionalResponses] we applied a quadratic penalty on the negative log-likelihood following:

```{r eval=FALSE, include=TRUE}
if(q < q_low){
  nll <- nll + penalty*(q-q_low)^2
} else{
  if(q >= q_up){
    nll <- nll + penalty*(q-q_up)^2
  } else{
    nll <- nll
  }
}
```

`q_low` is set to “0” by default (a type II functional response), `q_up` is set to “1” by default (a “strict” type III functional response), and `penalty` is set to 1000 by default. Especially `q_low` is important, as negative values may lead to an unsolvable time series for `gen_fr_sim` leading to a crash of the fitting process. Even with this restriction, the simulation may fail for extreme values of $F_{max}$ or $N_{half}$; in this case, the function returns `Inf`. Alternative solutions would be that the function returns `NA` [@Bolker2008EcologicalModelsData].

Required packages and their dependencies to be installed:

- `odin` [@FitzJohn2024OdinODEGeneration]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_parms_scan`

 **`gen_fr_parms_scan`** creates Latin hypercube samples for the functional response parameters in a reasonable range and calculates the according negative log-likelihood values. It returns the parameter values with the lowest negative log likelihood of these samples. Non-linear maximum likelihood fitting procedures require starting parameters, generally based on an educated guess [@Bolker2008EcologicalModelsData]. Moreover, these fits may end up in local best fits, and users should re-fit the data using different starting parameters [@Bolker2008EcologicalModelsData]. To overcome manually eyeballing as well as re-shuffling the starting parameters, @JagerAshauer2018ModellingSurvivalChemical suggested creating samples in a reasonable parameter range using and choosing the starting parameters (from the lowest nll value) from these samples. To reduce the number of required samples by keeping the variance of parameter values as wide as possible, it is recommended to use Latin hypercube sampling.  `gen_fr_parms_scan` requires the lhs package [@Carnell2024LhsLatinHypercubea].

Required packages and their dependencies to be installed:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]
- `odin` [@FitzJohn2024OdinODEGeneration]

Required packages to be attached:

- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]

#### `gen_fr_fit`

**`gen_fr_fit`** automatically fits the generalized functional response model [@Real1977KineticsFunctionalResponse; @RosenbaumRall2018FittingFunctionalResponses] to data. In the simplest case, you only need to provide the number of resources eaten (count data) and the initial resource density (count data): the function does the rest, including initial parameter value guessing. See the parameters section and the code example for more options. If your experiment ran a day, but you want to have the maximum feeding rate on an hourly basis, you can enter t_end = 24.

Required packages and their dependencies to be installed:

. `bbmle` [@BolkerEtAl2023BbmleToolsGeneral]
- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]
- `lhs` [@Carnell2024LhsLatinHypercubea]
- `odin` [@FitzJohn2024OdinODEGeneration]

Required packages to be attached:

- `dplyr` [@WickhamEtAl2023DplyrGrammarData]
- `foreach` [@MicrosoftWeston2022ForeachProvidesForeach]


## Funding Information

-   Mireia Aranbarri was funded by the **Investigo Programm funded by the NextGenerationEU initiative**.
-   Lorea Flores was funded by a grant by the **Spanish Ministry of Education and Culture**.
-   Ioar de Guzmán was funded by the **Spanish Ministry of Science, Innovation and Universities (TED2021-129966B-C31)**.
-   Julia Reiss was supported by a **Royal Society of London Starting Grant**.
-   Björn C. Rall gratefully acknowledges the funding by the **German Science Foundation (DFG) to the Research Unit DynaSym (FOR 5064)**.

## References
